---
title: "WILD 562 - Lab 2: Habitat USE"
author: "Mark Hebblewhite"
date: "1/16/2021"
output: github_document
---

```{r setup , include=FALSE}
require(knitr)
knitr::opts_chunk$set(echo = TRUE)
r <- getOption("repos")
r["CRAN"] <- "https://ftp.osuosl.org/pub/cran/"
options(repos = r)
```

# Lab 2 - Introduction to Analysis of Habitat Use by Wolves in Banff National Park

In the next 5 R computer labs (Labs 2 - 6), you have been charged with developing a conceptual framework for modeling wolf habitat use and selection to identify habitat for wolves as a function of a mix of environmental and 'biotic' variables, some of which we will explore today. These variables include elevation, landcover, human activity, and ungulate habitat suitability.  The next 4 (or more) labs will build on each other as we explore the definition of  wolf habitat using sample data from the Northern Rockies. In this first lab, you have been asked to describe wolf use of three habitat variables; elevation, ungulate species habitat suitability, and distance to human access.

## Ecological Background

Wolves (*Canis lupus*) are the most widely distributed terrestrial mammalian carnivore in the world, occurring in all continents except Africa and Antarctica, and all biomes from desert to tropical, arctic to grassland (Mech and Boitani 2003). Their only main requirement is the availability of large ungulate prey such as species of the Cervidae family. As such, wolves are habitat generalists, and not tied specifically to certain vegetation communities or types, and their densities where present are driven solely by ungulate biomass (Fuller and Sievert 2001), except when limited by human caused mortality. 

Wolves are persecuted by humans throughout their range because of direct conflict with livestock and because of perceptions (whether true or not) of competition with human hunters for large ungulate prey (Musiani and Paquet 2004). Throughout the history of wolf-human relationships, human persecution has been very successful, with wolves extirpated through much of their former range in North America and Europe especially during the 20th century. Wolf recovery, whether by recolonization via dispersal or active reintroduction such as in the Rocky Mountains of the USA (Fritts et al. 1997) is restoring wolves to ecosystems where they have been absent for decades or centuries, such as in Sweden. Yet, human persecution is still a leading cause of mortality for wolves, and can still limit their ability to recover to levels predicted by ungulate biomass (Thurber et al. 1994, Mladenoff et al. 1999, Fuller and Sievert 2001).

With the potential reintroduction of wolves to Colorado, biologists have been charged to define critical wolf habitat for wolves to guide recovery plans. Previous efforts to model wolf resource selection have used proxies for important drivers of wolf ‘habitat’ such as landcover type, elevation, etc., and have not fully considered human caused mortality as a ‘habitat’ factor in predicting wolf critical habitat. Moreover, scale dependence in habitat selection, and the definition of critical habitat, is a key concept which has been excluded from previous wolf habitat modeling, despite the crucial importance of scale (Johnson 1980, Boyce 2006).

Therefore, in the next 5 labs, you have been charged with developing a conceptual framework for modeling wolf habitat use and selection to define habitat for wolves as a function of the two key variables described in this short review; human activity and ungulate prey habitat suitability. The next 5 labs will build on each other as we explore the definition of wolf habitat using sample data from the Northern Rockies. In this first lab, you have been asked to estimate home ranges (Worton 1989, Seaman and Powell 1996, Powell 2000, Kie et al. 2010) that we will use to estimate availability in lab 3.  Second, we will describe wolf use of three main ecological variables; elevation, ungulate species habitat suitability, and distance to human access.

## Lab Objectives
The purposes of this lab are to:

1.	Practice preparation of spatial covariate data for ungulate biomass and human access information for spatial analyses.

2.	Learn how to make two kinds of home ranges (Minimum Convex Polygon and Fixed Kernel) to describe space use by wolves using the R package adeHabitatHR (Calenge et al). 

3.	Learn how to sample the availability of habitat using random sampling within animal home ranges. 

4.	Learn how to extract spatial covariate data from a landscape associated with used wolf telemetry locations.

5.	Learn how to conduct exploratory statistical analysis of the patterns of used data to draw conclusions about wolf use of habitat. 

## Data Descriptions

Data
*WOLFYHT* - We will investigate wolf resource selection in an area of the Northern Rockies during winters	of 2001-2005. Wolves were captured and radiocollared using foot-hold traps in summer, and 
helicopter netgunning and aerial darting during winter (U. of Alberta Animal care protocol ID# 
353112). We outfitted 14 wolves with VHF and 16 with GPS radiocollars (LMRT-3 or GPS3300sw, LOTEK Ltd.). All collared wolves were relocated aerially from a Cessna Skymaster 337 every 1-2 weeks, from the ground opportunistically. We will only use VHF telemetry location data for this lab collected during winter from two wolf packs; the Red Deer wolf pack and the Bow Valley wolf pack.

Landcover Covariates - We will use the following spatial databases to create the two main covariates 
related to wolf occupancy.

1.	**ELC_BNP** – previous field studies in BNP (Holroyd and Van Tighem 1983) developed Habitat Suitability Index Models based on counts of ungulate pellet densities and aerial survey data within different ELC polygons. (Hebblewhite et al. 2004) showed that wolf diet (i.e., 4th order habitat use) was composed of ~50% elk, 30% deer, 10% moose, and 10% other species such as bighorn sheep and mountain goats. We will test for wolf habitat selection for prey species by creating raster layers of each of the prey species as covered in Lab2.

2.	**HUMAN_ACCESS** – previous field research in the Canadian Rockies showed that wildlife species such as wolves avoided human activity (Whittington et al. 2004) and suffer increased mortality by humans near roads (Pletscher et al. 1997,Whittington et al. 2005a,Oakleaf et al. 2006). We will index vulnerability to human activity by creating a new variable, DISTANCE TO HUMAN ACCESS, following previous studies that showed wolves avoided areas of high human activity (Whittington et al. 2005).

3.	**ELEVATION** – ungulate prey and wolves are usually found in Valley bottoms during winter in the Northern Rockies because of snowfall at higher elevations, and we will therefore investigate the effects of elevation on wolf resource selection.

## Preliminaries: loading libraries

The R packages we will load up front today are:

```{r eval = TRUE, results = 'hide'}
packages <- c("ks", "plotrix", "lattice", "adehabitatHR", "maptools", "mapview", "raster", "ggplot2","colorRamps", "sf", "terra", "tmap", "stars", "dplyr")
```

Using the wrapper package reviewed in Lab 1:

```{r eval=TRUE, message=FALSE, results='hide'}

#function to install and load required packages
ipak <- function(pkg){
  new.pkg <- pkg[!(pkg %in% installed.packages()[, "Package"])]
  if (length(new.pkg)) 
    install.packages(new.pkg, dependencies = TRUE)
  sapply(pkg, require, character.only = TRUE)
}

#run function to install packages
ipak(packages)
```

# Objective 1: Managing Spatial Raster and Shapefile data

## Part 1a - Review

Let us review how to import shapefile datasets from ArcGIS shapefiles as a spatial data frame. Note because some of the subsequent calculations will take a long time, in the lab today I have provided some of the raster files, but demonstrate the code we used to calculate them. 

Also, as we gain familiarity with using spatial objects like shapefiles and rasters in R, I will continue to illustrate a few common analysis and data management approaches. As well as offer useful tutorials online for students new to using R for spatial data.  Here are 2

1) https://nceas.github.io/oss-lessons/spatial-data-gis-law/3-mon-intro-gis-in-r.html
and
2) https://www.neonscience.org/resources/learning-hub/tutorials/dc-plot-raster-data-r 

First we will read in shapefiles from the GISdata folder in your R project directory. 
```{r warning=FALSE}
elc_habitat<-st_read(here::here("Data","elc_habitat.shp"))
humanaccess<-st_read(here::here("Data","humanacess.shp"))
plot(elc_habitat) ## not sure if you want to keep this in here to show what base plot would do??
tmap_mode("plot") 
tm_shape(elc_habitat) + tm_sf()
tm_shape(humanaccess) + tm_sf()

wolfyht<-st_read(here::here("Data","wolfyht.shp"))
head(wolfyht)
```

As a preliminary step, lets plot some of the wolfyht using the base plot function, and using some of the structure of the wolfyht object. Lets remind ourselves of the structure of the SpatialPointsDataFrame first. 
```{r message=FALSE}
class(wolfyht)
crs(wolfyht,proj = TRUE) # note this is a UTM projected map system. 
str(wolfyht)
# Note that there are two fields, Easting and Northing which are the X and Y coordinates in UTM zone 11.  We will use these to map it for each PackID
# base plot of wolf packs by color with legend
base::plot(wolfyht$EASTING,wolfyht$NORTHING,col=c("red","blue")[wolfyht$PackID],ylab="Northing",xlab="Easting")
legend(555000,5742500,unique(wolfyht$Pack),col=c("blue","red"),pch=1) 
```

## Habitat Suitability Index Models
Next, we will explore/review some of the fields in the ELC database, lets make a plot of the Moose_winter ## habitat suitability index value for the study area. 

For background on how these 'rankings' of habitat suitability were created, see the associated readings here: 

Holroyd, G. L., and K. J. Van Tighem, editors. 1983. Ecological (biophysical) land classification of Banff and Jasper National Parks. Volume 3. The wildlife inventory. Canadian Wildlife Service, Edmonton.

Briefly, habitat suitability index models were developed by Holroyd and Van Tighem for Banff National Park using a variety of datasets, aerial surveys, pellet counts, and expert opinion. ELC polygons were ranked in order from low (0) to high (6), with lakes and water bodies classified as 7.  I will cover H.S.I. models, why you might want to build one, and how using an example for Bison in Banff later this semester. For now, lets continue. 

```{r warning=FALSE}
#For fun - if you want to play with using package tmap()
#construct tmap plot for Moose Winter habitat
tm_shape(elc_habitat)+tm_sf("MOOSE_W", border.alpha = 0)

#construct ggplot2 plot for Moose Winter Habitat
# Note - color = NA in the geom_sf() removes the border lines
elk_plot<-ggplot() + 
  geom_sf(data = elc_habitat, mapping = aes(fill = as.factor(MOOSE_W)), color = NA) + labs(x="Easting",y="Northing") + theme(axis.text.y = element_text(angle = 90, hjust=0.5))

#adjust fill colors of MOOSE_W  (note that I just selected some random colors, but made "7" as blue)
elk_plot2 <- elk_plot + scale_fill_manual(name="MOOSE_W",values=c("gray","gray", "red", "orange", "yellow", "green","darkblue"))
elk_plot2
```

Bighorn Sheep Winter Habitat Model
```{r}
#construct ggplot2 plot for Moose Winter Habitat
sheep_plot<-ggplot() + 
  geom_sf(data = elc_habitat, mapping = aes(fill = as.factor(SHEEP_W)), color = NA) + labs(x="Easting",y="Northing") + theme(axis.text.y = element_text(angle = 90, hjust=0.5))

#adjust fill colors of MOOSE_W  (note that I just selected some random colors, but made "7" as blue)
sheep_plot2 <- sheep_plot + scale_fill_manual(name="SHEEP_W",values=c("gray","gray", "red", "orange", "yellow", "green","darkblue"))
sheep_plot2
```

## Part 1b - Spatial Raster Operations

Here we will go through some typical raster operations when creating a set of spatially consistent raster layers to make into a 'stack' of rasters, a raster stack or brick, to then intersect animal locations with. 

First, we will use the fields for each of the Ungulate Habitat Suitability Index Models described in the associated readings from the ELC dataset.  We will then create a series of raster files, 'layers', for each species of winter ungulate in BNP using the `terra::rasterize()` command.  

The steps involved in creating these rasters below are key, and, good to review.  Remember, our goal is to create a consistent stack of rasters - a raster stack, or brick, of the same raster extent, resolution and projection for all subsequent analyses in R. 

1. Mask - we create a consistent raster MASK upon which to base all subsequent operations. This standardizes the extent, resolution and projection of the mask, and sets all values of the mask to be 0. 

2. `rasterize()` - we then use this operation to convert a vector file, a shapefile, to a raster. 

3. `resample()` - note that existing raster layers, e.g., elevation, distance to human access, had a different set of dimensions (extents) last week when we created our raster stack. They may have also had different resolutions, in which case we need to resample to a consistent resolution. 

```{r}
# Create a mask raster to use as a template for converting shapefile data to rasters
#create an empty raster
mask.raster <- rast()

#set extent (note that I customized this extent so it covered both elc_habitat and humanacess)
ext(mask.raster) <- c(xmin=443680.6, xmax=650430.4, ymin=5618405, ymax=5789236) 	

#set the resolution to 30 m 
res(mask.raster)<-30

#match projection to elc_habitat shapefile
crs(mask.raster)<- "+proj=utm +zone=11 +datum=NAD83 +units=m +no_defs +ellps=GRS80 +towgs84=0,0,0"

#set all values of mask.raster to zero
mask.raster[]<-0
```

Next I use the terra::rasterize() to create rasters for DEER_W, MOOSE_W, ELK_W, SHEEP_W, GOAT_W, WOLF_W 

```{r}
##This does not take as long as the raster() package and fasterize is no longer needed
deer_w<-terra::rasterize(elc_habitat, mask.raster, field="DEER_W")
moose_w<-terra::rasterize(elc_habitat, mask.raster, field="MOOSE_W")
elk_w<-terra::rasterize(elc_habitat, mask.raster, field="ELK_W")
sheep_w<-terra::rasterize(elc_habitat, mask.raster, field="SHEEP_W")
goat_w<-terra::rasterize(elc_habitat, mask.raster, field="GOAT_W")
wolf_w<-terra::rasterize(elc_habitat, mask.raster, field="WOLF_W")

#Alternative using Stars Package 
#deer_w <- st_as_stars(elc_habitat, name = attr(elc_habitat, "DEER_W"))
deer_w_stars <- st_rasterize(elc_habitat["DEER_W"])
moose_w_stars <- st_rasterize(elc_habitat["MOOSE_W"])
elk_w_stars <- st_rasterize(elc_habitat["ELK_W"])
sheep_w_stars <- st_rasterize(elc_habitat["SHEEP_W"])
goat_w_stars <- st_rasterize(elc_habitat["GOAT_W"])
wolf_w_stars <- st_rasterize(elc_habitat["WOLF_W"])
#plot result
plot(wolf_w)
plot(wolf_w_stars)
```

Next, we need to align the previously developed rasters, elevation and distance to human access, with the raster.mask developed above to have the exact same extent, spatial resolution, projection, etc.  We do so using the resample() function from the raster package. 

Learn more about resampling by typing `?terra::resample` 


```{r, eval = FALSE}
#resample elevation and humanaccess to match mask.raster 
#note - not sure how to do this with stars package yet -SMS - 1/24/2023
elevation2<-resample(elevation, mask.raster, method="bilinear")
disthumaccess2<-resample(disthumaccess, mask.raster, method="bilinear")
```

Exporting raster layers you have worked on using the `writeRaster` command from the terra package like this:
```{r, eval = FALSE}
#write raster layers to file
writeRaster(deer_w, here::here("Lab2","Output", "deer_w2.tiff"), overwrite = TRUE)
```

For today's lab we will us elevation rasters we created before the lab and re-load them from our working directory. 
```{r}
#reading in elevation raster with terra
elevation2<-rast(here::here("Data","Elevation2.tif")) #resampled

#reading in elevation raster with stars
elevation2_stars <- read_stars(here::here("Data","Elevation2.tif"))
```

How can we zoom into just one area of a raster object? There are many reasons we might want to, and many ways. One easy way is to 

```{r, warning = FALSE}
library(mapview)
mapview(wolfyht) + deer_w + sheep_w
```

## Part 1b - Distance to Raster (Human Access Layer)

First, there are many ways to evaluate human activity, polygonal (e.g., cities, parks), vector (roads, lines, points=houses), or more integrative spatial measures like density of human features or distance to human features.  Later in class we will compare and contrast why we would use distance to or density of a human feature as a resource covariate for wildlife. 

For now, I already created this raster layer because it took a LONG time on a desktop. So this might be an example where using ArcGIS might save time.But I include the code for illustration. 

```{r}
#first create an empty raster
dist.raster <- rast()
#set extent 
ext(dist.raster) <- ext(humanaccess)

#set the resolution to 30 m (Note that this takes a very long time with a 30 m resolution-even on my machine)
res(dist.raster)<-30

#match projection to humanaccess shapefile
crs(dist.raster)<- "+proj=utm +zone=11 +datum=NAD83 +units=m +no_defs +ellps=GRS80 +towgs84=0,0,0"

#set all values of dist.raster to zero
dist.raster[]<-0
```

Next, we need to 'rasterize` the new human access layer, and set human features to 1 using the dist.raster feature within rasterize. Note again, for time we will not actually run these today. 
```{r, eval = FALSE}
human.raster<-terra::rasterize(humanaccess, dist.raster, 1)

#calculate distance to human access- NOTE : DO NOT RUN THIS IT TAKES FOREVER; I ENDED UP DOING THIS CALCULATION JUST FOR THE POINTS
accessdist <-system.time(distance(human.raster))

```

Lets bring in the previously modeled distance to human layer. 
```{r}
disthumanaccess2<-rast(here::here("Data","DistFromHumanAccess2.tif")) 
plot(disthumanaccess2)
```

## Advanced question: how would you calculate distance to high human use?

In the human access shapefile, there is a field for the classification of human use on the trail from 0 to very high. We would then 'reclass' this field, and create a new shapefile. We could then use the steps above to rasterize this shapefile. 

```{r}
#first reclassify labels on humanaccess.shp file so they are correct (note: need to bring in humanaccess.shp above)
levels(as.factor(humanaccess$SUM_CLASS))
#[1] "0"         "High"      "HIGH"      "Low"       "LOW"      
#[6] "MEDIUM"    "Moderate"  "Nil"       "NIL"       "VERY HIGH"
#convert humanaccess$SUM_CLASS to a factor
humanaccess$SUM_CLASS<-as.factor(humanaccess$SUM_CLASS)

levels(humanaccess$SUM_CLASS)[1]<-"NIL"
levels(humanaccess$SUM_CLASS)[2]<-"HIGH"
levels(humanaccess$SUM_CLASS)[3]<-"LOW"
levels(humanaccess$SUM_CLASS)[5]<-"MODERATE"
levels(humanaccess$SUM_CLASS)[6]<-"NIL"

#create indicator variable for high or not high human access
highaccess<-humanaccess[humanaccess$SUM_CLASS=="HIGH" | humanaccess$SUM_CLASS=="VERY HIGH", ]

#use package tmap() to create simple plots
humanaccess_plot <- tm_shape(humanaccess)+tm_sf()
humanaccess_plot + tm_shape(highaccess) + tm_sf(col = "red") 
```
Now you would go through and rasterize this following the steps above to make this a raster file. Note I already did this to save time - lets load the distance to high human access layer. 
```{r}
#Reading in and plotting with terra
disthighhumanaccess<-rast(here::here("Data","DistFromHighHumanAccess2.tif"))
plot(disthighhumanaccess)

#reading in with stars
disthighhumanacess_stars <- read_stars(here::here("Data","DistFromHighHumanAccess2.tif"))
```

* Break Out Discussion *

# Objective 2 - Home Range Analysis 

Note I've done these here for the individual level I added the pack-level analysis below. But our goal will be to calculate 99% minimum convex polygon for individual wolves in the Red Deer wolf pack.  Our first step is to convert the SpatialPointsDataFrame to spatial points object. 

For these steps of estimating home ranges, we will be using the adehabitat set of packages created by Clement Calenge that are EXTREMELY valuable in habitat and spatial analyses in Ecology. 
 
Learn more here: 
 http://cran.nexr.com/web/packages/adehabitat/adehabitat.pdf
 
And in this paper: 
 Calenge, C. 2010. The package adehabitat for the R software: a tool for the analysis of space and habitat use by animals. Ecological Modeling 197:516-519. https://doi.org/10.1016/j.ecolmodel.2006.03.017 
 
First we will do a bit of data management to extract just the basics we need from the wolf telemetry object, wolfyht.shp.  We will extract the X Y locations, define a new dataframe, add the XY coordinates, a projection and create a new SpatialPointsDataFrame file - a common set of steps that here is slightly unnecessary because we already have a spatial points data frame.  But useful to review especially for students creating a new spatialPointsDataFrame from say, an excel spreadsheet or data from GPS radiocollars. 

```{r warning=FALSE}
rd.data<-wolfyht[wolfyht$Pack=="Red Deer",]
x<-rd.data$EASTING
y<-rd.data$NORTHING
xy<-cbind(x,y)
class(xy)
rd <- data.frame(as.character(rd.data$NAME))
coordinates(rd) <- xy
crs(rd) <-  "+proj=utm +zone=11 +datum=NAD83 +units=m +no_defs +ellps=GRS80 +towgs84=0,0,0"

class(rd)
# Fit 99% mpc
#cp.rd <- mcp(rd, percent=99)
#note error that one animal does not have at least 5 locations
table(rd.data$NAME)
#42 60 69 70 81 82 84 
#43 25  4 15  3  2  1  
#looks like 4 of the wolves do not have enough locations

rd.test <- st_drop_geometry(rd.data) %>% 
  filter(!NAME %in% c("69","81","82","84"))


#remove these individuals with too few of locations
names(rd)<-"NAME"
rd<-rd[rd@data$NAME!="69" & rd@data$NAME!="81" & rd@data$NAME!="82" & rd@data$NAME!="84",]
#remove unused NAME levels
rd@data$NAME<-factor(rd@data$NAME)
```

Next we will fit a 99% Minimum Convex Polygon to these cleaned data for EACH individual wolf. 
```{r, warning= FALSE}
# Fit 99% mcp
cp.rd <- mcp(rd, percent=99)
plot(rd, col="black")
plot(cp.rd[cp.rd@data$id=="42",], col="blue", add=TRUE)
plot(cp.rd[cp.rd@data$id=="70",], col="green", add=TRUE)
plot(cp.rd[cp.rd@data$id=="60",], col="red", add=TRUE)
plot(rd, col="black", add=TRUE)

#check area in square meters for each Red Deer wolf pack
as.data.frame(cp.rd)

#calculate area for different percents of MPC in square meters. 
mcp.area(rd, percent=seq(50, 100, by=5))
```

We will now estimate Kernel Density Estimate home ranges using the command `kernelUD()`.
```{r, warning= FALSE}
#calculate 99% KDE for Red Deer wolf pack
red.deerUD <- kernelUD(rd, grid=30, extent=0.5, same4all=TRUE) # reference grid
image(red.deerUD)

#get polygons for home ranges
homerangeRD <- getverticeshr(red.deerUD)
as.data.frame(homerangeRD)
class(homerangeRD)
plot(homerangeRD, col=2:4)

#Estimate UD in raster mode
red.deerud <- getvolumeUD(red.deerUD) 
red.deerud

## Set up graphical parameters for the output of getvolumeUD 
par(mar=c(0,0,2,0)) #set margin
image(red.deerud[[1]]) #for first wolf only
title("Red Deer Wolf UD") 
xyzv <- as.image.SpatialGridDataFrame(red.deerud[[1]]) 
contour(xyzv, add=TRUE)
```

Next we will store the volume under the UD (as computed by getvolumeUD) 
of the first animal in fud as a RASTER file of 1= home range, 0 = not home range. 
```{r, warning= FALSE}
fud <- red.deerud[[1]] #for first wolf only
## store the value of the volume under UD in a vector hr95 
hr95 <- as.data.frame(fud)[,1] 
## if hr95 is <= 95 then the pixel belongs to the home range
## (takes the value 1, 0 otherwise)
hr95 <- as.numeric(hr95 <= 95) 
## Converts into a data frame 
hr95 <- data.frame(hr95) 
## Converts to a SpatialPixelsDataFrame 
coordinates(hr95) <- coordinates(red.deerud[[1]])
gridded(hr95) <- TRUE 
## display the results 
image(hr95)
```

## Bow Valley wolves
Calculate 99% minimum convex polygon for individual wolves in Bow valley wolf pack repeating the same steps as above. 
```{r, warning= FALSE}
#first convert the spatialpointsdataframe to spatial points object
bv.data<-wolfyht[wolfyht$Pack=="Bow valley",]
x<-bv.data$EASTING
y<-bv.data$NORTHING
xy<-cbind(x,y)

bv <- data.frame(as.character(bv.data$NAME))
coordinates(bv) <- xy
crs(bv) <-  "+proj=utm +zone=11 +datum=NAD83 +units=m +no_defs +ellps=GRS80 +towgs84=0,0,0"

# Fit 99% mpc
cp.bow <- mcp(bv, percent=99)
plot(bv, col="black")
plot(cp.bow[cp.bow@data$id=="63",], col="blue",add=TRUE)
plot(cp.bow[cp.bow@data$id=="87",], col="red",add=TRUE,)
plot(cp.bow[cp.bow@data$id=="44",], col="green",add=TRUE)
plot(bv, col="black", add=TRUE)
```

## Red Deer wolves
```{r, warning= FALSE}
#check area for each Red Deer wolf pack
as.data.frame(cp.bow)

#calculate area for different percents of MPC
mcp.area(bv, percent=seq(50, 100, by=5))

#calculate 99% KDE for Red Deer wolf pack
bow.valleyUD <- kernelUD(bv, grid=30, extent=0.1, same4all=TRUE) # reference grid
image(bow.valleyUD)
```
```{r}
#get polygons for home ranges
homerangeBV <- getverticeshr(bow.valleyUD)
as.data.frame(homerangeBV)
class(homerangeBV)
plot(homerangeBV, col=2:4)

#Estimate UD in raster mode
bow.valleyud <- getvolumeUD(bow.valleyUD) 
bow.valleyud

## Set up graphical parameters for the output of getvolumeUD 
par(mar=c(0,0,2,0)) #set margin
image(bow.valleyud[[1]])
title("Bow Valley Pack UD") 
xyzv <- as.image.SpatialGridDataFrame(bow.valleyud[[1]]) 
contour(xyzv, add=TRUE)
```

Store the volume under the UD (as computed by getvolumeUD) of the first animal in fud. 
```{r}
fud <- bow.valleyud[[1]]
## store the value of the volume under 95% UD in a vector hr95 
hr95 <- as.data.frame(fud)[,1] 
## if hr95 is <= 95 then the pixel belongs to the home range
## (takes the value 1, 0 otherwise)
hr95 <- as.numeric(hr95 <= 95) 
## Converts into a data frame 
hr95 <- data.frame(hr95) 
## Converts to a SpatialPixelsDataFrame 
coordinates(hr95) <- coordinates(bow.valleyud[[1]])
gridded(hr95) <- TRUE 
## display the results 
image(hr95)
```

## Calculate 99% minimum convex polygon for both wolf packs
```{r, warning=FALSE}
#first convert the spatialpointsdataframe to spatial points object
x<-wolfyht$EASTING
y<-wolfyht$NORTHING
xy<-cbind(x,y)

all <- data.frame(as.character(wolfyht$Pack))
coordinates(all) <- xy
crs(all) <-  "+proj=utm +zone=11 +datum=NAD83 +units=m +no_defs +ellps=GRS80 +towgs84=0,0,0"

# Fit 99% mpc
cp.all <- mcp(all, percent=99)

plot(wolfyht, col="black")
plot(cp.all[cp.all@data$id=="Bow valley",], col="blue",add=TRUE)
plot(cp.all[cp.all@data$id=="Red Deer",], col="green",add=TRUE)
plot(wolfyht, col="black", add=TRUE)
```

Checking areas, etc 
```{r, warnings = FALSE}
#check area for each Red Deer wolf pack
as.data.frame(cp.all)

#calculate area for different percents of MPC
mcp.area(all, percent=seq(50, 100, by=5))

#calculate 99% KDE for both wolf packs
allUD <- kernelUD(all, grid=30, extent=0.5, same4all=TRUE) # reference grid
image(allUD)
```

Get polygons for home ranges
```{r}
homerangeALL <- getverticeshr(allUD)
as.data.frame(homerangeALL)
class(homerangeALL)
plot(homerangeALL, col=2:3)
```

Estimate UD in raster mode
```{r}
allud <- getvolumeUD(allUD) 
allud

## Set up graphical parameters for the output of getvolumeUD 
par(mar=c(0,0,2,0)) #set margin
image(allud[[1]]) #for first wolf only
title("Output of getvolumeUD") 
xyzv <- as.image.SpatialGridDataFrame(allud[[1]]) 
contour(xyzv, add=TRUE)
```

Store the volume under the UD (as computed by getvolumeUD) of the first animal in fud 
```{r}
fud <- allud[[1]] #for first wolf pack only
## store the value of the volume under UD in a vector hr95 
hr95 <- as.data.frame(fud)[,1] 
## if hr95 is <= 95 then the pixel belongs to the home range
## (takes the value 1, 0 otherwise)
hr95 <- as.numeric(hr95 <= 95) 
## Converts into a data frame 
hr95 <- data.frame(hr95) 
## Converts to a SpatialPixelsDataFrame 
coordinates(hr95) <- coordinates(allud[[1]])
gridded(hr95) <- TRUE 
## display the results 
image(hr95)
```

# Objective 3 - Sampling Availability Within Home Ranges 

The critical step in most used-available designs for habitat models, including so-called presence-only models, is the sampling of availability. We are going to dive right in without much care to the conceptual foundation for sampling availability in any way, and just do it here today with the home ranges we created for each MCP wolf pack above. 

However, as an exercise, we will consider one wrinkle in sampling availability randomly within our Kernel home ranges. Recall our pixel resolution was 30m. It might be 'interesting' to consider the problem of whether we want to sample the exact same spatial unit for an 'available' sample. One way to do this is ensure that no randomly generated points are within 30m of each other. 
```{r}
#subset polygons by wolf pack
red.deerPOLY<-homerangeALL[homerangeALL@data$id=="Red Deer",]
bow.valleyPOLY<-homerangeALL[homerangeALL@data$id=="Bow valley",]

#generate 1000 points from Red Deer wolf pack KDE polygon
rd.avail<-spsample(red.deerPOLY, 1000, "random")
plot(rd.avail)

#generate 1000 points from Bow valley wolf pack KDE polygon
bv.avail<-spsample(bow.valleyPOLY, 1000, "random")
plot(bv.avail)
```

Lets plot them all together, used and home-range level availability. 
```{r}
plot(wolfyht$EASTING,wolfyht$NORTHING, col=c("red","blue")[wolfyht$PackID],ylab="Northing",xlab="Easting")
legend(555000,5742500,unique(wolfyht$Pack),col=c("blue","red"),pch=1)
plot(bv.avail, add=TRUE)
plot(rd.avail, add=TRUE)
```

**Discuss in class: assumptions of sampling availability this way? Spatial scale of this sampling design?**

